Conditional Power and Friends: The Why and How of (Un)planned, Unblinded Sample Size Recalculations in Confirmatory Trials

### 2020-10-13 latex/manuscript.tex
Updates from Overleaf

% other
\usepackage{color}
\usepackage[numbers,super,sort&compress]{NJDnatbib}
\bibliographystyle{WileyNJD-AMA}
\usepackage{scalerel}
\newcommand{\kr}[1]{{\color{red}#1}}

@@

\author{
  Kevin Kunzmann \thanks{
        MRC Biostatistics Unit, University of Cambridge,
        Cambridge Institute of Public Health,
        Forvie Site, Robinson Way,
        Cambridge Biomedical Campus,
        Cambridge CB2 0SR,
        United Kingdom
    }\\
    MRC Biostatistics Unit \\
    University of Cambridge \\
    \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk} \\
    Michael J. Grayling \\
    Population Health Sciences Institute \\
    Newcastle University \\
    \texttt{michael.grayling@newcastle.ac.uk} \\
    Kim May Lee \\
    Pragmatic Clinical Trials Unit\\
    Queen Mary University of London\\
    \texttt{k.m.lee@qmul.ac.uk} \\
    David S.\ Robertson \\
    MRC Biostatisics Unit\\
    University of Cambridge\\
    \texttt{david.robertson@mrc-bsu.cam.ac.uk} \\
    Kaspar Rufibach \\
    Methods, Collaboration, and Outreach Group (MCO) \\
    Department of Biostatistics\\
    F. Hoffmann-La Roche, Basel\\
    \texttt{kaspar.rufibach@roche.com} \\
    James M. S. Wason \\
    Population Health Sciences Institute \\
    Newcastle University\\
    and \\
     MRC Biostatistics Unit \\
    University of Cambridge \\
    \texttt{james.wason@newcastle.ac.uk}
}

@@

\begin{abstract}
Determining the sample size of a confirmatory trial is challenging.
If the sample size is too small, the chance of detecting an effect is small.
If it is too big, trial duration and costs are needlessly inflated.
Adapting the final sample size of a trial to the evidence accruing
during the trial is thus a natural way to address
this planning uncertainty.
In a regulatory environment,
designs with adaptive sample size need to
account for their optional stopping to guarantee
strict type-I error-rate control.
A variety of different methods to maintain type-I error-rate control
after unplanned changes of the initial sample size have been proposed
in the literature rendering interim analyses for
the purpose of sample size recalculation feasible in
practice.
Any form of interim analysis allows the analysis of
a trial's prospects to finally reject the null hypothesis
given the data observed up to the interim analysis.
Since the sample size is usually determined via an
argument based on the power of the trial,
an interim analysis raises the question of
how the final sample size should be determined conditional on the accrued information.
Conditional power is a concept often put forward in this context.
Since it depends on the unknown effect size,
we take a strict estimation perspective and compare
assumed conditional power, observed conditional power,
and predictive power with respect to their properties as estimators
of the unknown conditional power.
We then demonstrate that pre-planning an interim analysis using
methodology for unplanned interim analyses is ineffective and
naturally leads to the concept of optimal two-stage designs.
We conclude that unplanned design adaptations should only be conducted
as reaction to trial-external new evidence, operational needs to violate the originally chosen design,
or \textit{post~hoc} changes in the objective criterion.
Finally, we show that commonly discussed sample size recalculation rules can lead to paradoxical outcomes and propose two alternative ways of reacting to
newly emerging trial-external evidence.
\end{abstract}

@@

The three estimators are depicted in Figure~\ref{fig:acp-ocp-pp-comparison-absolute} as functions
of the observed interim outcome, where the interim time-point $m = 26$.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_example}
        \caption{%
            Estimates as function of the interim data.
        }
        \label{fig:acp-ocp-pp-comparison-absolute}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_sampling_distributions}
        \caption{%
            Histograms of the sampling distributions.
        }
        \label{fig:sampling-distributions}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_bias_mae_mse}
        \caption{%
            Bias, mean absolute error (MAE) and mean squared error (MSE).
        }
        \label{fig:acp-ocp-pp-as-estimators}
    \end{subfigure}
    \caption{%
        Properties of $\ACP,\OCP$, and $\PP$ as estimators of the unknown conditional power at $m=26$ an overall sample size $n=79$.
    }
\end{figure}
The most sensitive measure is $\OCP$ while $\ACP$ is the least sensitive
(see also spread of the sampling distributions in Figure~\ref{fig:sampling-distributions}).
This is due to $\ACP$ assuming a fixed effect size and thus
only being affected by direct changes in $z_m$.
Both $\OCP$ and $\PP$, however, are also indirectly affected by updating
the belief about the effect size with the interim results.
The plug-in estimate $\OCP$ assumes that the observed effect is the
true effect and does not quantify the uncertainty around this value in any way.
$\PP$, on the other hand, invokes Bayes' theorem and then integrates
$\CP(z_m,c,\theta)$ with respect to the obtained posterior distribution.
The degree of adaptation of $\PP$ thus depends
on the vagueness of the chosen prior.
In fact, as the prior approaches a point mass at $\theta_1$, $\PP$ converges
to $\ACP$.
$\ACP$ can thus be seen as a mere special case of $\PP$ with a point prior
on $\theta_1>0$.

@@

We illustrate both approaches by example of the optimal two-stage design that minimizes the objective
$\E_\varphi[\,n(Z_m)\,]$ subject to a maximal type-I error rate of $0.025$ and a minimal expected power of $0.8$, see~Figure~\ref{fig:binding-recalculation}.
This design has a pre-planned interim analysis after $m=35$ individuals.
Now assume that
the prior $\varphi$ needs to be revised due to new trial-external
information at the pre-planned interim analysis.
We consider members of the class of shifted truncated Normal distributions
\begin{align}
    \psi_{\mu}(\theta) := \frac{\displaystyle \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\phi\left(\frac{\theta - \mu}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - \mu}{0.2}\right) - \Phi\left(\frac{-0.5 - \mu}{0.2}\right)}
\end{align}
as revised priors, i.e. the original prior~$\varphi=\psi_{0.4}$.
Figure~\ref{fig:recalculation-old-m} shows the behavior of the recalculated sample size and critical value in this scenario for observed effects of $z_m/\sqrt{m}=0.25, 0.3, 0.35$ and varying mean of the revised prior if the originally planned interim time-point is maintained.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/recalculation-type-II-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect.
        }
        \label{fig:recalculation-old-m}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/recalculation-type-II-earlier}
        \caption{%
            Earlier-than-planned interim analysis at an observed effect of $0.3$ and varying interim time-point.
        }
        \label{fig:recalculation-earlier}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/recalculation-lambda-approach-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect and the $\lambda$ approach.
        }
        \label{fig:recalculation-lambda}
    \end{subfigure}
    \caption{%
        Comparison of consistent unplanned recalculation schemes.
    }
\end{figure}
Clearly, the original design is indeed invariant if the prior remains unchanged.
The sample size is increased if the new prior is more conservative
than $\varphi$ and \textit{vice~versa}.

>>>>

% other
\usepackage{color}
\usepackage[numbers,super,sort&compress]{NJDnatbib}
\bibliographystyle{WileyNJD-AMA}
\usepackage{scalerel}
\newcommand{\kr}[1]{{\color{red}#1}}
\usepackage{comment}

@@

%\begin{comment}
\author{
    Kevin Kunzmann \thanks{
        MRC Biostatistics Unit, University of Cambridge,
        Cambridge Institute of Public Health,
        Forvie Site, Robinson Way,
        Cambridge Biomedical Campus,
        Cambridge CB2 0SR,
        United Kingdom,
        \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk}
    }\\
    MRC Biostatistics Unit \\
    University of Cambridge \\
    Michael J. Grayling \\
    Population Health Sciences Institute \\
    Newcastle University \\
    Kim May Lee \\
    Pragmatic Clinical Trials Unit\\
    Queen Mary University of London\\
    David S.\ Robertson \\
    MRC Biostatisics Unit\\
    University of Cambridge\\
    Kaspar Rufibach \\
    Methods, Collaboration, and Outreach Group (MCO) \\
    Department of Biostatistics\\
    F. Hoffmann-La Roche, Basel\\
    James M. S. Wason \\
    Population Health Sciences Institute \\
    Newcastle University\\
    and \\
     MRC Biostatistics Unit \\
    University of Cambridge
}
%\end{comment}
%\author{}

@@

\begin{abstract}
Adapting the final sample size of a trial to the evidence accruing
during the trial is a natural way to address planning uncertainty.
Designs with adaptive sample size need to
account for their optional stopping to guarantee
strict type-I error-rate control.
A variety of different methods to maintain type-I error-rate control
after unplanned changes of the initial sample size have been proposed
in the literature.
This makes interim analyses for the purpose of sample size 
recalculation feasible in a regulatory context.
Since the sample size is usually determined via an
argument based on the power of the trial,
an interim analysis raises the question of
how the final sample size should be determined conditional on the accrued information.
Conditional power is a concept often put forward in this context.
Since it depends on the unknown effect size,
we take a strict estimation perspective and compare
assumed conditional power, observed conditional power,
and predictive power with respect to their properties as estimators
of the unknown conditional power.
We then demonstrate that pre-planning an interim analysis using
methodology for unplanned interim analyses is ineffective and
naturally leads to the concept of optimal two-stage designs.
We conclude that unplanned design adaptations should only be conducted
as reaction to trial-external new evidence, operational needs to violate the originally chosen design,
or \textit{post~hoc} changes in the objective criterion.
Finally, we show that commonly discussed sample size recalculation rules can lead to paradoxical outcomes and propose two alternative ways of reacting to
newly emerging trial-external evidence.
\end{abstract}

@@

The three estimators are depicted in Figure~\ref{fig:acp-ocp-pp-comparison-absolute} as functions
of the observed interim outcome, where the interim time-point $m = 26$.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_example}
        \caption{%
            Estimates as function of the interim data.
        }
        \label{fig:acp-ocp-pp-comparison-absolute}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_sampling_distributions}
        \caption{%
            Histograms of the sampling distributions.
        }
        \label{fig:sampling-distributions}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_bias_mae_mse}
        \caption{%
            Bias, mean absolute error (MAE) and mean squared error (MSE).
        }
        \label{fig:acp-ocp-pp-as-estimators}
    \end{subfigure}
    \caption{%
        Properties of $\ACP,\OCP$, and $\PP$ as estimators of the unknown conditional power at $m=26$ an overall sample size $n=79$.
    }
\end{figure}
The most sensitive measure is $\OCP$ while $\ACP$ is the least sensitive
(see also spread of the sampling distributions in Figure~\ref{fig:sampling-distributions}).
This is due to $\ACP$ assuming a fixed effect size and thus
only being affected by direct changes in $z_m$.
Both $\OCP$ and $\PP$, however, are also indirectly affected by updating
the belief about the effect size with the interim results.
The plug-in estimate $\OCP$ assumes that the observed effect is the
true effect and does not quantify the uncertainty around this value in any way.
$\PP$, on the other hand, invokes Bayes' theorem and then integrates
$\CP(z_m,c,\theta)$ with respect to the obtained posterior distribution.
The degree of adaptation of $\PP$ thus depends
on the vagueness of the chosen prior.
In fact, as the prior approaches a point mass at $\theta_1$, $\PP$ converges
to $\ACP$.
$\ACP$ can thus be seen as a mere special case of $\PP$ with a point prior
on $\theta_1>0$.

@@

We illustrate both approaches by example of the optimal two-stage design that minimizes the objective
$\E_\varphi[\,n(Z_m)\,]$ subject to a maximal type-I error rate of $0.025$ and a minimal expected power of $0.8$, see~Figure~\ref{fig:binding-recalculation}.
This design has a pre-planned interim analysis after $m=35$ individuals.
Now assume that
the prior $\varphi$ needs to be revised due to new trial-external
information at the pre-planned interim analysis.
We consider members of the class of shifted truncated Normal distributions
\begin{align}
    \psi_{\mu}(\theta) := \frac{\displaystyle \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\phi\left(\frac{\theta - \mu}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - \mu}{0.2}\right) - \Phi\left(\frac{-0.5 - \mu}{0.2}\right)}
\end{align}
as revised priors, i.e. the original prior~$\varphi=\psi_{0.4}$.
Figure~\ref{fig:recalculation-old-m} shows the behavior of the recalculated sample size and critical value in this scenario for observed effects of $z_m/\sqrt{m}=0.25, 0.3, 0.35$ and varying mean of the revised prior if the originally planned interim time-point is maintained.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figures/recalculation-type-II-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect.
        }
        \label{fig:recalculation-old-m}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figures/recalculation-type-II-earlier}
        \caption{%
            Earlier-than-planned interim analysis at an observed effect of $0.3$ and varying interim time-point.
        }
        \label{fig:recalculation-earlier}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figures/recalculation-lambda-approach-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect and the $\lambda$ approach.
        }
        \label{fig:recalculation-lambda}
    \end{subfigure}
    \caption{%
        Comparison of consistent unplanned recalculation schemes.
    }
\end{figure}
Clearly, the original design is indeed invariant if the prior remains unchanged.
The sample size is increased if the new prior is more conservative
than $\varphi$ and \textit{vice~versa}.
